`make.modifiers`

### 2020-12-02 inst/gitbook/03_make.modifiers.Rmd
Update 03_make.modifiers.Rmd

@@

In the examples above, we have seen how to specify modifications to the birth death process (via `branch.length`, `selection` and `speciation`), however, one might note that these modifications are not dynamic.
In other words, throughout the process, the modifications remain constant (even if they are conditional).
It is however possible to code the modifiers

In the following section we will see how to program these functions in a `"modifiers"` before learning about other aspects of the birth death process that can be modified (the `selection` and `speciation` steps).

As shown in the example with the `branch.length` modifier, it is possible to modify a specific part of the birth death process (e.g. the change in time) but in the previous example, the `constant.brlen` modifier only included a constant modification of the process whereas it is probably more valuable to have a variable change in branch length (e.g. like the default `branch.length` modifier that is dependent on the number of taxa at the time of the simulation).
It is possible add conditions and modifications to the `"modifiers"` object through the `condition` and `modify` function.
As stated in their names, the `condition` function is indicating in which conditions to do a special action while the `modify` function is indicating what to modify (typically when a condition is met).

`condition` and `modify` are hard coded in the `branch.length` function that they concern, i.e. they are variables (functions) within the function.

```{r, eval = FALSE}
constant.brlen <- function(bd.params      = NULL,
                           n.taxa         = NULL,
                           parent.lineage = NULL,
                           trait.values   = NULL,
                           modify.fun     = NULL) {
    return(1)
}

constant_brlen <- dads(bd.params = list(extinction = 0.25),
                     stop.rule = list(max.living = 20),
                     traits    = make.traits(),
                     modifiers = make.modifiers(
                        branch.length = constant.brlen))

plot(constant_brlen, main = "constant branch length")
```

```{r, eval = FALSE}
random.brlen <- function(bd.params = NULL,
                         n.taxa = NULL,
                         parent.lineage = NULL,
                         trait.values = NULL,
                         modify.fun = NULL) {
    return(runif(1))
}

random.spec  <- function(bd.params = NULL,
                         n.taxa = NULL,
                         parent.lineage = NULL,
                         trait.values = NULL,
                         modify.fun = NULL) {
    return(sample(c(TRUE, FALSE), 1))
}

set.seed(1)
yule_tree <- dads(bd.params = list(extinction = 0),
                     stop.rule = list(max.living = 20),
                     traits    = make.traits(),
                     modifiers = make.modifiers(
                        branch.length = random.brlen,
                        speciation    = random.spec))

plot(yule_tree, main = "a (failing) yule(ish) tree")
```

## The condition function (`condition`)

## The modify function (`modify`)

>>>>

### The allowed arguments {#allowarguments}

It is of course possible to design some more advanced function to interact with the birth death process.
You can create a function for `branch.length`, `selection` and `speciation` that involve any of the following arguments:

  * `bd.params`: a named list containing `"numeric"` values that contains the birth death parameters (at least `"speciation"` and `"extinction"`);
  * `lineage`: a named list containing the lineage data (see below).
  * `trait.values`: a `"matrix"` containing `"numeric"` values with the trait names as column names and the lineages ID as row numbers (you can use it with the function `parent.traits` to access the trait of the previous node for example).
  * `modify.fun`: a `"list"` of named `"function"` (usually passed through `condition` and `modify`).

The `lineage` list contains the following elements (missing elements are allowed):

 * `lineage$parents`: an `"integer"` vector: the list of parent lineages;
 * `lineage$livings`: an `"integer"` vector: the list of lineages still not extinct;
 * `lineage$drawn`: a single `"integer"`: the ID of the selected lineage;
 * `lineage$current`: a single `"integer"`: the selected lineage (is equal to `lineage$livings[lineage$drawn]`);
 * `lineage$n`: a single `"integer"`: the current number of non extinct lineage (is equal to `length(lineage$livings))`;
 * `lineage$split`: a `"logical"` vector: the list of splits for each lineage (`TRUE`), the number of total tips is equal to `sum(!lineage$split)`.

In general, unless you know what you're doing, you can ignore most arguments for specific modifiers since they are handled automatically within the `dads` function.
Therefore any argument can be left undeclared or missing and is always handled internally. For example, if you did not declare `n.taxa` as a function argument but are using `n.taxa` in the function, `n.taxa` will be detected and treated as a current argument automatically as set accordingly within the birth death process (e.g. `n.taxa` will be set to the current number of taxa every iteration of the process).

For example, we can create a function that increases branch length proportional to the number of species "alive" at each time of the simulation in a discrete way. I.e. for discrete numbers of taxa, the branch length increases by jumps (ten fold):

```{r}
## A more complex binned.branch.length function
increasing.brlen <- function(bd.params, lineage) {

    ## Setting the cumulated birth and death
    birth_death <- bd.params$speciation + bd.params$extinction

    ## Returning branch lengths depending on different number of taxa
    if(lineage$n <= 5) {
        return(1    * rexp(1, sum(5 * birth_death)))
    }
    if(lineage$n <= 10) {
        return(10   * rexp(1, sum(10 * birth_death)))   
    }
    if(lineage$n <= 15) {
        return(100  * rexp(1, sum(15 * birth_death)))      
    }
    if(lineage$n <= 20) {
        return(1000 * rexp(1, sum(20 * birth_death)))
    } else {
        return(1000 * rexp(1, sum(lineage$n * birth_death)))
    }
}
```

We can then create it as a `"modifiers"` object and run a new simulation:

```{r}
## Creating a modifiers
increasing_modifier <- make.modifiers(branch.length = increasing.brlen)

## Generating a new tree with this modifier
set.seed(0)
increasing_tree <- dads(bd.params = bd_params,
                        stop.rule = stope_rule,
                        modifiers = increasing_modifier)
```

And we can visualise the difference between the resulting trees:

```{r}
par(mfrow = c(1,3))
plot(default_tree,    main = "Default modifier")
plot(modified_tree,   main = "Constant branch length\nmodifier")
plot(increasing_tree, main = "Increasing branch length\nmodifier (binned)")
```

## The selection function (`selection`)

The `selection` function is used in the birth death process to know which lineage to select when running a speciation (or extinction!) event.
By default, this function randomly selects one taxon that is currently not extinct (using: `sample(1:n.taxa, 1))`.
Similarly as `branch.length` it is possible to modify this part of the birth death process.
For example, we could simply select always the last created lineage (to create a "ladder" or most asymmetric tree):

```{r}
## Our function to always select the last taxon
## (making sure it returns an integer)
select.last <- function(lineage) {
    return(as.integer(lineage$n))
}
```

> Note that here the function can only intake the [allowed arguments as described above](#allowarguments) (here `n.taxa`: the number of current living taxa).

We can then create a `"modifiers"` object the same way as before using this time the `selection` argument:

```{r}
## A modifier for selection
ladderised_modifier <- make.modifiers(selection = select.last)
## Generating a new tree with this modifier
set.seed(0)
ladderised_tree <- dads(bd.params = bd_params,
                        stop.rule = stope_rule,
                        modifiers = ladderised_modifier)
## Displaying the results
par(mfrow = c(1,2))
plot(default_tree,    main = "Default modifier")
plot(ladderised_tree, main = "Ladderising modifier")
```

Again, it is of course possible to make the modifier more complex and in combination with other elements of the tree.
For example, we can create a `"dads"` object that also creates a trait a add to it a `selection` modifier that only selects for tips with positive trait values.

```{r}
## Our function that only select taxa with positive trait values
select.positive <- function(trait.values, lineage) {
    ## Selecting the taxa with positive traits only
    positives <- as.integer(rownames(trait.values)[which(trait.values[, 1] >= 0)])

    ## Select the current taxa that descend from a node with a positive value
    positive_living <- cbind(lineage$parents, seq_along(lineage$split)
                             )[which(lineage$parents %in% positives), 2]

    ## Select one tip randomly in the ones with descendants with positive values
    return(sample(which(lineage$livings %in% positive_living), 1))
}

## Creating the modifier
positive_skew <- make.modifiers(selection = select.positive)

## Creating a (default) trait object
BM_trait <- make.traits()

## Simulate a tree and trait with no modifier
set.seed(1)
default_dads <- dads(bd.params = bd_params,
                     stop.rule = stope_rule,
                     traits    = BM_trait)

## Simulate a tree and trait with the modifier
set.seed(1)
skewed_trait_dads <- dads(bd.params = bd_params,
                          stop.rule = stope_rule,
                          traits    = BM_trait,
                          modifiers = positive_skew)

## Plotting the differences in trees and traits
par(mfrow = c(1, 2))
plot(default_dads, main = "Default trait and tree")
plot(skewed_trait_dads, main = "Skewed trait and tree")
```

The third function that can be used to modify the birth death process is the `speciation` function.
This one is used during the birth death process to decide whether a lineages speciates (creating a node and two new lineages) or goes extinct (creating a tip).

> Note that the `speciation` function only affects tips or nodes before the simulation reaches the `stop.rule`.
The then surviving lineages are all automatically transformed into tips.

By default, the `speciation` function is trigger a speciation even if a number randomly drawn from a uniform distribution is lower than the ratio between the speciation and the speciation and extinction parameter. If the randomly drawn number is higher, the lineage goes extinct.

```{r, eval = FALSE}
## The speciation in pseudo-code:
runif(1) < speciation/ (speciation + extinction)
```

Creating a `"modifiers"` with a `speciation` function works the same way as for `branch.length` and `selection` but the function that will be used needs to output a logical value (see table [below](#summarymodifiers)).
Once the function is created simply input your function for speciation in the modifier and run the `dads` function with that modifier:

```{r, eval = FALSE}
## Speciating or going extinct randomly
## (regardless of the extinction parameter)
random.extinct  <- function() {
    return(sample(c(TRUE, FALSE), 1))
}

## Creating the modifiers object
random_extinction <- make.modifiers(speciation = random.extinct)

## Generating a new tree with this modifier
set.seed(0)
modified_tree <- dads(bd.params = bd_params,
                      stop.rule = stope_rule,
                      modifiers = random_extinction)

par(mfrow = c(1,2))
plot(default_tree,  main = "Default modifier")
plot(modified_tree, main = "Random extinction\nmodifier")
```

Note how every lineage end up going extinct!
And again, we can make some more advanced modifiers: for example, one where a tip always goes extinct if their ancestor has a negative trait value (here we will also introduce the utility function `parent.trait` that automatically selects the trait values of the parent of the current lineage.

```{r}
## A modifier for removing tips with negative values
bias.trait <- function(trait.values, lineage) {
    if(parent.traits(trait.values, lineage) < 0) {
        ## Go extinct!
        return(FALSE)
    } else {
        ## Speciate
        return(TRUE)
    }
}

## Creating the modifier
biased_trait <- make.modifiers(speciation = bias.trait)

## Simulate a tree and trait with the modifier
set.seed(1)
biased_trait_dads <- dads(bd.params = bd_params,
                          stop.rule = stope_rule,
                          traits    = BM_trait,
                          modifiers = biased_trait)

## Plotting the differences in trees and traits
par(mfrow = c(1, 2))
plot(default_dads, main = "Default trait and tree")
plot(biased_trait_dads, main = "Biased trait and tree")
```

## Summary of the inputs and outputs for the `branch.length`, `selection` and `speciation` modifiers {#summarymodifiers}

modifier name   | accepted input (arguments) | required output (class) |
----------------|----------------------------|-------------------------|
`branch.length` | `bd.params`, `lineage`, `trait.values` | `"numeric"` |
`selection`     | `bd.params`, `lineage`, `trait.values` | `"integer"` |
`speciation`    | `bd.params`, `lineage`, `trait.values` | `"logical"` |

@@

In the examples above, we have seen how to specify modifications to the birth death process (via `branch.length`, `selection` and `speciation`), however, one might note that these modifications are not dynamic.
In other words, throughout the process, the modifications remain constant (even if they are conditional).
It is however possible to code the `"modifiers"` so that they can be affected by `"events"` objects (see next chapter).

To do so, you can formally declare conditions (`condition`) and modifications (`modify`) as internal functions that can then be modified my an `"events"` object.
`condition` and `modify` are hard coded in the `branch.length` function that they concern, i.e. they are variables (functions) within the function.
